---
title: "IDS intestigation worksheet"
author: "by Kate Supremacy: Connor, Kate, Ruth, Torgua & Kate"
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, message = FALSE}
library(tidyverse)
# Add any other libraries here
library(lubridate)


```


```{r load-data}
# load your data 
nyweather <- read_csv("data/ny_weather.csv")

nyweather <- nyweather %>%
  rename(
    "date" = `DATE`,
    "cool_deg_hrs" = `HLY-CLDH-NORMAL`,
    "cloud_brkn_pct" = `HLY-CLOD-PCTBKN`,
    "cloud_clr_pct" = `HLY-CLOD-PCTCLR`,
    "cloud_few_pct" = `HLY-CLOD-PCTFEW`,
    "cloud_ovrcst_pct" = `HLY-CLOD-PCTOVC`,
    "cloud_scat_pct" = `HLY-CLOD-PCTSCT`,
    "dewp_10" = `HLY-DEWP-10PCTL`,
    "dewp_90" = `HLY-DEWP-90PCTL`,
    "dewp_avg" = `HLY-DEWP-NORMAL`,
    "heat_index_avg" = `HLY-HIDX-NORMAL`,
    "heat_deg_hrs" = `HLY-HTDH-NORMAL`,
    "pres_10pctl" = "HLY-PRES-10PCTL",
    "pres_90pctl" = "HLY-PRES-90PCTL",
    "pres_avg" = `HLY-PRES-NORMAL`,
    "temp_10pctl" = "HLY-TEMP-10PCTL",
    "temp_90ptcl" = "HLY-TEMP-90PCTL",
    "temp_avg" = `HLY-TEMP-NORMAL`,
    "wind_chill_avg" = `HLY-WCHL-NORMAL`,
    "wind_main_dir" = "HLY-WIND-1STDIR",
    "wind_main_pct" = "HLY-WIND-1STPCT",
    "wind_2nd_dir" = "HLY-WIND-2NDDIR",
    "wind_2nd_pct" = "HLY-WIND-2NDPCT",
    "wind_spd_avg" = `HLY-WIND-AVGSPD`,
    "pct_calm" = `HLY-WIND-PCTCLM`,
    "wind_vector_dir" = `HLY-WIND-VCTDIR`,
    "wind_vector_spd" = `HLY-WIND-VCTSPD`
  ) %>% # renamed columns
  select(-c("STATION", "NAME")) %>% # got rid of station and name as we know these
  separate(date, into = c("day", "time"), sep = "T") # separated into day and time (might want to change variable types)
#please also note that pressure will not change!! so is not worth looking at



```
```{r}
# Which direction is the prevailing wind most frequently in? 
nyweather %>%
  count(wind_main_dir) %>%
  arrange(desc(n)) 
# Using the frequency table and the document in the reference tab of the Google doc, the prevailing wind most frequently in the North West direction.

# Which direction is the secondary wind most frequently in?
nyweather %>%
  count(wind_2nd_dir) %>%
  arrange(desc(n))
# Using the frequency table and the document in the reference tab of the Google doc, the secondary wind most frequently in the West direction.

# Both the prevailing wind and the secondary winds least frequent direction is East.

```
```{r}
#Creating a month variable so different weather features can be compared over the year
nyweather_separate <- nyweather %>% 
  separate(day, into = c("Month", "Day"), sep = "-") %>% 
  mutate(
    Month = as.numeric(Month),
    Day = as.numeric(Day),
    Name_Month = factor(month.abb[Month], levels = month.abb)
  )
```


```{r}
nyweather_monthly <- nyweather_separate %>% 
  group_by(Name_Month) %>% 
  summarize(monthly_temp_avg= mean(temp_avg),
            monthly_dewp_avg= mean(dewp_avg),
            monthly_wind_chill_avg = mean(wind_chill_avg),
            monthly_wind_spd_avg = mean(wind_spd_avg))

  
nyweather_monthly %>% 
ggplot(aes(x = Name_Month, y= monthly_temp_avg))+
  geom_bar(stat = "identity") +
  labs(title = "Average Temperature Per Month in New York",
       x = "Month",
       y= "Temperature in °F")

# This shows what can be assumed, The summer months are hotter on average than the winter months

nyweather_monthly %>% 
ggplot(aes(x = Name_Month, y= monthly_wind_spd_avg))+
  geom_bar(stat = "identity") +
  labs(title = "Average Wind Speed Per Month in New York",
       x = "Month",
       y= "Wind Speed in Mph")
#This shows that October to April is the windier period of the year, with the summer months having a smaller windspeed. I don't actually know if it's in Miles per Hour but id guess so

nyweather_monthly %>% 
ggplot(aes(x = Name_Month, y= monthly_dewp_avg))+
  geom_bar(stat = "identity") +
  labs(title = "Average Dew Point Per Month in New York",
       x = "Month",
       y= "Dew Point in °F")

#The dew point follows a similar curve to that of temperature, suggesting a relationship between the two, from research the higher the dew point the more humid it feels, long periods of a high dew point implies prolonged humidity.

nyweather_monthly %>% 
ggplot(aes(x = Name_Month, y= monthly_wind_chill_avg))+
  geom_bar(stat = "identity") +
  labs(title = "Average Wind Chill Per Month in New York",
       x = "Month",
       y= "Wind Chill in °F")
#This also follows a similiar curve as temperature, the wind chill is simply the temperature that the wind makes it, so for example in summer the wind chill is not going to be very active so the temperature is very close to the same as the inital recording but in the colder months the wind chill can make the temperature feel 10 degrees colder.

#These last two will be particularly clear when compared directly with the temperature graph as a decent amount of analysis can be derived from the comparison. Such as a large gap between the temperature and the wind chill temperature implies periods of very cold winds.
```