---
title: "New York Weather Investigation"
subtitle: "Investigating the relationship between wind and other weather variables"
author: "Kate Supremacy <br> Connor, Kate, Ruth, Torgua and Kate"
institute: "University of Edinburgh"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      
---

```{r install-package, include = FALSE, eval = FALSE}
# Copy and paste the following code into your console to download and install
# the `xaringan` package that contains all of the code which allows you 
# to create presentation slides in Rmarkdown
install.packages('xaringan')
```


```{r load-packages, include = FALSE}
# Add any additional packages you need to this chunk
library(tidyverse)
library(tidymodels)
library(palmerpenguins)
library(knitr)
library(xaringanthemer)
library(xaringanExtra)
library(patchwork)
use_xaringan_extra("panelset")
```

```{r setup, include=FALSE}
# For better figure resolution
knitr::opts_chunk$set(fig.retina = 3, dpi = 300, fig.width = 6, fig.asp = 0.618, out.width = "80%")
```

```{r load-data, include=FALSE}
nyweather <- read_csv("data/ny_weather.csv")

nyweather_unclean <- nyweather

nyweather <- nyweather %>%
  rename(
    "date" = `DATE`,
    "cool_deg_hrs" = `HLY-CLDH-NORMAL`,
    "cloud_brkn_pct" = `HLY-CLOD-PCTBKN`,
    "cloud_clr_pct" = `HLY-CLOD-PCTCLR`,
    "cloud_few_pct" = `HLY-CLOD-PCTFEW`,
    "cloud_ovrcst_pct" = `HLY-CLOD-PCTOVC`,
    "cloud_scat_pct" = `HLY-CLOD-PCTSCT`,
    "dewp_10" = `HLY-DEWP-10PCTL`,
    "dewp_90" = `HLY-DEWP-90PCTL`,
    "dewp_avg" = `HLY-DEWP-NORMAL`,
    "heat_index_avg" = `HLY-HIDX-NORMAL`,
    "heat_deg_hrs" = `HLY-HTDH-NORMAL`,
    "pres_10pctl" = "HLY-PRES-10PCTL",
    "pres_90pctl" = "HLY-PRES-90PCTL",
    "pres_avg" = `HLY-PRES-NORMAL`,
    "temp_10pctl" = "HLY-TEMP-10PCTL",
    "temp_90pctl" = "HLY-TEMP-90PCTL",
    "temp_avg" = `HLY-TEMP-NORMAL`,
    "wind_chill_avg" = `HLY-WCHL-NORMAL`,
    "wind_main_dir" = "HLY-WIND-1STDIR",
    "wind_main_pct" = "HLY-WIND-1STPCT",
    "wind_2nd_dir" = "HLY-WIND-2NDDIR",
    "wind_2nd_pct" = "HLY-WIND-2NDPCT",
    "wind_spd_avg" = `HLY-WIND-AVGSPD`,
    "pct_calm" = `HLY-WIND-PCTCLM`,
    "wind_vector_dir" = `HLY-WIND-VCTDIR`,
    "wind_vector_spd" = `HLY-WIND-VCTSPD`
  ) %>% # renamed columns
  select(-c("STATION", "NAME")) %>% # got rid of station and name as we know these
  separate(date, into = c("day", "time"), sep = "T") # separated into day and time (might want to change variable types)
#please also note that pressure will not change!! so is not worth looking at

#replacing missing data with nas
nyweather[nyweather == -777.7] <- NA

nyweather_pres <- nyweather %>% 
  select(-cool_deg_hrs,-pres_10pctl,-pres_90pctl,-pres_avg,-temp_10pctl,-temp_90pctl,-dewp_10,-dewp_90)
```

```{r include=FALSE}

#Background image
style_xaringan(
  title_slide_background_image = "img/new_york_city_night_cityscape.jpg"
)
```

class: center, middle

## Investigating the relationship between wind and other weather variables

---

class: inverse, center, middle

# Breakdown of the Dataset 

---

# Ny Weather

```{r data, include=TRUE, echo=FALSE}
nyweather_pres %>% 
  head()
```

- These are just some of the variables in the data set, with the initial set containing 28 columns

- Most of the variable names are fairly easy to pick up and some of the less obvious ones will be explained later

---

class: center, middle

# The columns can be split into 4 key categories

```{r, echo=FALSE}
library(tibble)

# Create the table
categories <- data.frame(
  "Cloud analysis" = c("cloud_brkn_pct, cloud_clr_pct, cloud_few_pct, cloud_ovrcst_pct, cloud_scat_pct"),
  "Temperature analysis" = c("temp_avg, dewp_avg, heat_index_avg, wind_chill_avg"),
  "Wind analysis" = c("wind_main_dir, wind_main_pct, wind_2nd_dir, wind_2nd_pct, wind_speed_avg, wind_vector_dir, wind_vector_spd"),
  "Time periods" =c("day, time"))
  


# Display the table
knitr::kable(categories, align = "c", col.names = c("Cloud Analysis", "Temperature Analysis", "Wind Analysis", "Time Periods"))
```
---
# Cleaning The Data
.panelset[
.panel[.panel-name[Renaming Columns]
The initial data set named variables using a different and confusing notation, to combat this we had to rename all but 1 of the 28 variables.
```{r Rename, results='hide'}
  nyweather_unclean %>%
  rename(
    "date" = "DATE",
    "cool_deg_hrs" = "HLY-CLDH-NORMAL",
    "cloud_brkn_pct" = "HLY-CLOD-PCTBKN",
    "cloud_clr_pct" = "HLY-CLOD-PCTCLR",
    "cloud_few_pct" = "HLY-CLOD-PCTFEW",
    "cloud_ovrcst_pct" = "HLY-CLOD-PCTOVC",
    "cloud_scat_pct" = "HLY-CLOD-PCTSCT")
```
]
.panel[.panel-name[Dealing with NA values]

The NA values were also of a different format.
```{r NA values, results='hide'}
nyweather[nyweather == -777.7] <- NA
```


[*]See Citations
]
.panel[.panel-name[Conversion of Temperature Units]
As the measurements were made in New York all temperatures were measured in Fahrenheit
```{r Temp Units, results='hide'}
nyweather <- nyweather %>% 
  mutate(across(c(dewp_10,dewp_90,dewp_avg,
                  heat_index_avg,temp_10pctl,
                  temp_90pctl,temp_avg,wind_chill_avg),
                ~ signif((5/9)*(.x-32), 3)))
```
]
.panel[.panel-name[Percentage Correction]
All the columns with percentage values were a factor of 10 larger than they should have been. 
```{r Pct Correction, results='hide'}
nyweather <- nyweather %>% 
  mutate(across(c(cloud_brkn_pct,cloud_clr_pct,
                  cloud_few_pct,cloud_ovrcst_pct,
                  cloud_scat_pct, wind_main_pct,wind_2nd_pct),
                ~ .x/10))
```
]
]
---
class: inverse, center, middle
# Exploratory Data Analysis 

---

###Frequency Tables 

What are the most frequent prevailing wind directions?
.pull-left[
Main Wind Direction
```{r Freq Table 1, echo=FALSE}

freq_table_1 <- nyweather %>%
  count(wind_main_dir) %>%
  arrange(desc(n)) 

freq_table_1

```
]
.pull-right[
2nd Wind Direction
```{r Freq Table 2, echo=FALSE}
freq_table_2 <- nyweather %>%
  count(wind_2nd_dir) %>%
  arrange(desc(n))

freq_table_2
```
]
---
###Histograms
Histograms were made of many variables including these key few that are reused in other data analysis
```{r Histos, echo=FALSE}
#Temperature
plot_1 <- ggplot(data=nyweather,
       mapping=aes(
         x= temp_avg))+
  geom_histogram(
    binwidth = 0.7
  )
#Heat index
plot_2 <- ggplot(data=nyweather,
       mapping=aes(x=heat_index_avg))+
  geom_histogram(
    binwidth = 1
  )
#Wind Chill
plot_3 <- ggplot(data=nyweather,
       mapping=aes(x=wind_chill_avg))+
  geom_histogram(
    binwidth=1
  )
plot_4 <- ggplot(data=nyweather,
       mapping=aes(x=wind_vector_dir))+
  geom_histogram(
    binwidth=3
  )

  plot_1 + plot_2 + plot_3 + plot_4 

```

---
### Bar Charts
.panelset[
.panel[.panel-name[Data set Prep]
```{r Bar Code}
nyweather_separate <- nyweather %>% 
  separate(day, into = c("Month", "Day"), sep = "-") %>% 
  mutate(
    Month = as.numeric(Month),
    Day = as.numeric(Day),
    Name_Month = factor(month.abb[Month], levels = month.abb)
  )

nyweather_monthly <- nyweather_separate %>% 
  group_by(Name_Month) %>% 
  summarize(monthly_temp_avg= mean(temp_avg),
            monthly_dewp_avg= mean(dewp_avg),
            monthly_wind_chill_avg = mean(wind_chill_avg),
            monthly_wind_spd_avg = mean(wind_spd_avg))

```
]
.panel[.panel-name[Charts]

```{r Charts, echo=FALSE}
bar_1 <- nyweather_monthly %>% 
ggplot(aes(x = Name_Month, y= monthly_temp_avg))+
  geom_bar(stat = "identity") +
  labs(title = "Monthly Average Temperature",
       x="",
       y= "Temperature in °C")+
  theme(axis.text.x = element_text(angle = 45))

bar_2 <- nyweather_monthly %>% 
ggplot(aes(x = Name_Month, y= monthly_wind_spd_avg))+
  geom_bar(stat = "identity") +
  labs(title = "Monthly Average Wind Speed",
       x="",
       y= "Wind Speed in Mph")+
  theme(axis.text.x = element_text(angle = 45))

bar_3 <- nyweather_monthly %>% 
ggplot(aes(x = Name_Month, y= monthly_dewp_avg))+
  geom_bar(stat = "identity") +
  labs(title = "Monthly Average Dew Point",
       x="",
       y= "Dew Point in °C")+
  theme(axis.text.x = element_text(angle = 45))

bar_4 <- nyweather_monthly %>% 
ggplot(aes(x = Name_Month, y= monthly_wind_chill_avg))+
  geom_bar(stat = "identity") +
  labs(title = "Monthly Average Wind Chill",
       x="",
       y= "Wind Chill in °C")+
  theme(axis.text.x = element_text(angle = 45))

bar_1 + bar_2 + bar_3 + bar_4
```
]]
---
###Scatterplots

In an attempt to find relationships between variables we made scatter plots.
Examples:

- Heat Index v Wind Chill 

- Wind Speed v Wind Chill 

- Dew Point v Wind Chill

- Dew Point v Heat Index
---
class: center
```{r, echo=FALSE}
ggplot(nyweather, aes(x = wind_chill_avg, y = heat_index_avg)) +
  geom_jitter() + 
  labs(title = "Heat Index Avg V Wind Chill Avg")

```
---
class: center
```{r, echo=FALSE}
ggplot(nyweather, aes(x = wind_spd_avg, y = wind_chill_avg)) +
  geom_point() + 
  labs(title = "Wind Chill Avg v Wind Speed Avg") 

```
---
class: center
```{r, echo=FALSE}
ggplot(nyweather, aes(x = wind_chill_avg, y = dewp_avg)) +
  geom_jitter() +
  labs(title = "Dew Point Avg v Wind Chill Avg")
```
---
class: center
```{r, echo=FALSE}
ggplot(nyweather, aes(x = heat_index_avg, y = dewp_avg)) +
  geom_point() +
  labs(title = "Dew Point Avg V Heat Index Avg")
```
---
class: center, inverse, middle
#Linear Regression Model
---
#Heat Index and Wind Chill
Given the apparent relationship shown by the scatter plot we investigated further.
.panelset[.panel[.panel-name[Creating the Model]
```{r}
set.seed(1234)
nyweather_split2 <- initial_split(nyweather)
nyweather_train2 <- training(nyweather_split2)
nyweather_test2  <- testing(nyweather_split2)

wind_heat_mod <- linear_reg() %>%
  set_engine("lm")

wind_heat_rec <- recipe(
  wind_chill_avg ~ heat_index_avg,
  data = nyweather_train2)

wind_heat_wflow <- workflow() %>%
  add_model(wind_heat_mod) %>%
  add_recipe(wind_heat_rec)

wind_heat_fit <- wind_heat_wflow %>%
  fit(data = nyweather_train2)
```
]
.panel[.panel-name[Applying Test Data]
```{r}
wind_heat_pred <- predict(wind_heat_fit, nyweather_test2) %>%
  bind_cols(nyweather_test2 %>% select(wind_chill_avg, heat_index_avg))
```

```{r}
rsq(wind_heat_pred, truth = wind_chill_avg, estimate = .pred)
rmse(wind_heat_pred, truth = wind_chill_avg, estimate = .pred)
```
]
.panel[.panel-name[Analysing Outputs]

.pull-left[
-The R squared value being 0.994 implies the majority of the variation in points is explained by the variables chosen

-But such a high R squared value can be a case of over fitting
]
.pull-right[
-The Rmse value of 0.793 implies that the predictions made will be off by around 0.793 units

-The value isn't very low so also implies average prediction accuracy
]

When these are both taken into account it appears that the model is successful, as it can predict fairly well and the variables explain most of the data.
]
.panel[.panel-name[Residual Plot]
```{r, echo=FALSE, include=FALSE}
invisible({wind_heat_fit <- wind_heat_wflow %>%
  fit(data = nyweather_train2)
wind_heat_fit_tidy <- tidy(wind_heat_fit) 
wind_heat_fit_aug  <- augment(wind_heat_fit, nyweather_train2) %>%
  mutate(res_cat = ifelse(.resid > 0, TRUE, FALSE))
wind_heat_fit_aug}) # Created an augmented table with the residuals.
```
.pull-left[
-Unfortunately when plotting the residuals there is a clear pattern in plots

-This suggests that the actual relationship is nonlinear despite earlier impressions

-Upon reflection a linear model was not a suitable choice as it fails to demonstrate the entirety of the relationship between the variables
]
.pull-right[
```{r,  echo=FALSE}
ggplot(wind_heat_fit_aug, mapping = aes(x = .pred, y = .resid)) +
  geom_point(alpha = 0.4) + 
  geom_hline(yintercept = 0, 
             color = "blue", 
             lty = "dashed") +
  labs(title = "Residual Plot", 
       x = "Predicted Wind Chill", 
       y = "Residuals")  
```
]]]
---
class: center, inverse, middle
#Logistical Regression Model
---
#Predicting Wind Direction From Temperature

.panelset[.panel[.panel-name[Preparing Data]
```{r}
nyweather <- nyweather %>%
  mutate(
    wind_main_west = case_when(
      wind_main_dir == "6" ~ "yes",
      wind_main_dir == "7" ~ "yes",
      wind_main_dir == "8" ~ "yes",
      TRUE ~ "no"
    ),
    wind_main_west = fct_relevel(wind_main_west, "no", "yes")
  )
```
Creating a binary variable is necessary for a logistical regression model hence we have made a variable that states if the wind is travelling west or not.
]
.panel[.panel-name[Creating the Model]
```{r}
set.seed(1234)
nyweather_split <- initial_split(nyweather)
nyweather_train <- training(nyweather_split)
nyweather_test  <- testing(nyweather_split) # Splitting data into training and testing.

wind_dir_rec_1 <- recipe(
  wind_main_west ~ temp_avg,
  data = nyweather_train)  # Created a recipe.

wind_dir_mod_1 <- logistic_reg() %>%
  set_engine("glm")     # Created a model.

wind_dir_wflow_1 <- workflow() %>%
  add_model(wind_dir_mod_1) %>%
  add_recipe(wind_dir_rec_1) #Added them to a work flow.
```
]
.panel[.panel-name[Equation of the Line]
```{r}
wind_dir_fit_1 <- wind_dir_wflow_1 %>%
  fit(data = nyweather_train)  # Applying work flow to training data.

tidy(wind_dir_fit_1)
```
You can see from the table below, that as the average temperature increases the smaller the logic of the probability, and therefore the smaller the probability that the wind will be in the west direction. Or in other words if it is colder it is more likely the wind will be going in the west direction (Including north west and south west direction).
$$
\log\left(\frac{p}{1-p}\right) = 2.63 - 0.17 \cdot \text{temp_avg}
$$
]
.panel[.panel-name[ROC Curve]
```{r, echo=FALSE, results='hide'}
wind_dir_pred_1 <- predict(wind_dir_fit_1, nyweather_test, type = "prob") %>% 
  bind_cols(nyweather_test) 
wind_dir_pred_1
```

.pull-left[
```{r}
wind_dir_pred_1 %>%
  roc_curve(
    truth = wind_main_west,
    .pred_yes,             
    event_level = "second") %>%
  autoplot() # Plotting a ROC curve
```
]
.pull-right[
```{r}
wind_dir_pred_1 %>%
  roc_auc(
    truth = wind_main_west,
    .pred_yes,             
    event_level = "second"
  ) 
```
]]]
---
#Multivariate Prediction

.panelset[.panel[.panel-name[Creating a New Model]
```{r}
wind_dir_rec_2 <- recipe(
  wind_main_west ~ temp_avg + heat_index_avg + wind_chill_avg,
  data = nyweather_train  # Created a recipe.
)
wind_dir_rec_2 <- wind_dir_rec_2 %>%
  step_dummy(all_nominal(), -all_outcomes()) 
# Created dummy variables for all predictor variables and not of outcome variables.

wind_dir_mod_2 <- logistic_reg() %>%
  set_engine("glm")     # Created a model.

wind_dir_wflow_2 <- workflow() %>%
  add_model(wind_dir_mod_2) %>%
  add_recipe(wind_dir_rec_2) #Added them to a work flow.
```
]
.panel[.panel-name[Equation of the Line]
```{r}
wind_dir_fit_2 <- wind_dir_wflow_2 %>%
  fit(data = nyweather_train)  # Applying work flow to training data.

tidy(wind_dir_fit_2) 
```
The interpretation is slightly more complex, interestingly by adding variables, on average the mean temperature has the opposite effect on the probability of the wind direction.
$$
\log\left(\frac{p}{1-p}\right) = -5.26 + 7.77 \cdot \text{temp_avg} - 5.57 \cdot \text{heat_index_avg} - 1.95 \cdot \text{wind_chill_avg}
$$
]
.panel[.panel-name[ROC Curve]
```{r, echo=FALSE, results='hide'}
wind_dir_pred_2 <- predict(wind_dir_fit_2, nyweather_test, type = "prob") %>% 
  bind_cols(nyweather_test) 
wind_dir_pred_2 # Using fitted model to predict the testing data.
```
.pull-left[
```{r}
wind_dir_pred_2 %>%
  roc_curve(
    truth = wind_main_west,
    .pred_yes,             
    event_level = "second") %>%
  autoplot() # Plotting a ROC curve
```
]
.pull-right[
```{r}
wind_dir_pred_2 %>%
  roc_auc(
    truth = wind_main_west,
    .pred_yes,             
    event_level = "second") 
```
]]]
---
class: center, middle, inverse
#Answering our question

---
#Reflection
.pull-left[
What Went Well

- We made successful data visualisations while trying to keep the 4 respects in mind.

- We worked well as a team, had an organised investigation document with comments to allow for pair programming techniques.

- We used a google doc that detailed all meetings, tasks and plans which allowed us all to work productively and stay on task.

]
.pull-right[
Areas for Improvement

- We could have attempted a further investigation into the cloud scatterings and how they change with wind direction or speed.

- We could have tried cross validation as a way to verify our model

- If we had imported another data set containing precipitation records
]
---
#Citations
- Arguez, A., Durre, I., Applequist, S., Squires, M., Vose, R., Yin, X. and Bilotta, R. (2010). NOAA's U.S. Climate Normals (1981-2010): JFK International Airport, NY US, 2010. Available at: https://www.ncei.noaa.gov/access/metadata/landing-page/bin/iso?id=gov.noaa.ncdc:C00824 (Accessed 11 November 2024).

- Pedersen, T.L. (n.d.) 'patchwork'. Available at: https://patchwork.data-imaginist.com/ (Accessed 23 November 2024)

- Pix4free (2016). Free New york city night cityscape Image. Available at: https://pix4free.org/assets/library/2021-01-12/originals/new_york_city_night_cityscape.jpg (Accessed 21 November 2024). 
